Continuous Deductive Reasoning over Streamed RDF Data

### 2015-12-01 article.tex
Complete rewrite of most of the introduction to neaten and bring more to the point.  May include HORRORS

  \begin{abstract}
    In this paper, we introduce Continuous Datalog, an extension of
    conventional Datalog that adds the concept of streams of
    temporally-annotated facts to conventional Datalog and defines the
    notion of entailment over a windowed stream. We go on to define CoDeR,
    an abstract model for continuous deductive reasoning that provides
    both a data stream model based on Continuous Datalog and a minimal
    algebra for expressing queries and programs.  Finally, we describe the
    R4 stream reasoner, which implements the CoDeR model as a Rete-based
    dataflow network and provide a brief comparative evaluation of R4
    against ETALIS and Sparkwave.

@@

\begin{nestedsection}{Introduction}{intro}
  % general framing of the problem (why reason over streams)
  As the Semantic Web becomes more extensive, with more people providing and describing their data in the semantic formats recommended by the W3C, the variety in the nature of the data expressed continues to approach that of previous, more established data formats.
  This includes the growing number of use cases for the continuous processing of streams of semantic data, in which large volumes of time-sensitive data must be processed in near-real time, with the aim of minimising the volume of data transported, duplicated and stored across the Web whilst maximising the variety of problems that can be solved.
  With the focus on reasoning in the Semantic Web stack, varying in expressivity from the low complexity RDF entailment rules through to the high complexity OWL Full, it becomes clear that the greatest challenge in semantic stream processing is that of stream reasoning.
  In the most recent literature, this has been characterised as a layer of higher complexity processing on top of Complex Event Processing (CEP) \citep{mileo15webSR}, deeming it suitable only for application to low-throughput macro-event streams.
  However, Semantic Web reasoning languages provide many varying degrees of expressivity corresponding to to varying levels of complexity.
  Furthermore, in many cases static ontologies are applicable to streamed data (such as the description and classification of readings from sensor networks, for example), and therefore do not require the features unique to CEP (and temporal reasoning \citep{OrgunWadge92,Tuzhilin93}), rather requiring a concept of data becoming less relevant over time.
  A prime expression of low-complexity reasoning that is capable of such tasks in a static context is the rule language profile of OWL 2 (OWL 2 RL) \citep{w3cowl2profiles}.
  The continuous application of reasoning of this nature to streamed RDF data may obviously be characterised as `stream reasoning', but occupies the same general complexity bracket as stream querying in the hierarchy presented by \citep{mileo15webSR}, making it applicable to high-throughput streams.
  It is this low-complexity continuous reasoning that is the focus of this paper.

  The value of such processing is the same as that of reasoning over comparatively static data, that being that the division of knowledge into particular knowledge and general knowledge allows for the expression of a greater volume of information in a lesser volume of data.
  In addition, this division allows the application of a single domain of general knowledge to a variety of distinct sets of particular knowledge, and the application of many distinct domains of general knowledge to a single set of particular knowledge.
  It should be noted that this division of general and particular knowledge in the expression of a reasoning problem depends on the generalisation of that problem:
  in Description Logics, it may be considered that the TBox of a knowledge base describes general knowledge and the ABox describes particular knowledge;
  however, the fundamental rules by which entailments are derived from knowledge bases by Description Logics constitute a more general knowledge than that of the TBox of some specific knowledge base, by which the entirety of a knowledge base may be considered particular knowledge.
  In the latter case, a distinction may still be made between a TBox and ABox in the context of streaming, with the former being expressed in static data and the latter being streamed or vice versa, though this is dependent on the rate at which the TBox and ABox are expected to change.
  In the existing stream processing literature, the vast majority of current use cases consider the application of some \emph{specific} static domain of general knowledge (e.g. a specific TBox) to some stream of particular knowledge;
  thus, with stream reasoning the volume of streamed data may be dramatically reduced by extracting the static domain of general knowledge that underpins the data and providing that domain as a static ontology, then streaming only the minimal particular knowledge that would be necessary to deduce the full data set.

  % motivating scenario
  Take, for example, the case of extensive sensor networks with many and varied sensors, such as those served by the SemsorGrid4Env project\footnote{\url{http://www.semsorgrid4env.eu/}}.
  Any stand-alone semantic representation of this data may contain classification information for each sensor, according to what they measure, as well as how frequently and how accurately, as well as other factors.
  It may also contain explicit data concerning events that are reported by no one sensor, and are instead extrapolated from the base sensor data.
  In both of these cases, the classification criteria and extrapolation process for macro-events may be expressed as general knowledge that is applicable to all sensor readings.
  As such, a consumer of the streams produced using SemsorGrid4Env could download the appropriate ontologies once, then use them to reason over a reduced set of streamed data containing only the base readings from the sensor network, thus obtaining the classification and macro-event information desired without having to stream it explicitly.
  Furthermore, sensor readings are often superseded by later readings of the same type from the same sensor, and so older readings may be discarded as time progresses.

  % *brief* summary of state of the art (to be expanded in related work)
  Recent research has approached the task of stream reasoning from two different directions.
  %   RDF stream processing and reasoning
  The RDF stream processing community has proposed extensions to their continuous semantic query languages to support the \emph{iterative/recursive execution} of queries over the results of other queries \citep{C-SPARQL,EP-SPARQL,walavalkar08streamingkb}, in which each query expresses a rule in a logic program.
  This reflects precisely the approach to reasoning using Datalog in deductive databases, in which queries may be executed on intensional relations whose definitions must be used in turn to query the database, recursing in this manner until base only extensional relations are queried.
  Meanwhile, the reasoning community has explored the concept of continuously reasoning over streamed data in a variety of forms, primarily through a range of truth maintenance algorithms \citep{dred,inc-materialisation,stream-truth-maintenance}.
  %%%FIXME
  NEED SOMETHING ON TIME-DECAY MODEL

  %   issues with existing art: no explicit semantics for reasoning
  While the iterative execution of queries as rules is sufficient for performing reasoning to the expressivity of Datalog when applying one-shot queries to static data, as practised in deductive databases, the query-by-query windowing semantics of most stream-querying systems do not translate to a coherent semantics for multi-rule stream reasoning of the kind investigated in this paper.
  This is because the windowing semantics proposed for stream querying involve applying the entirety of a ``problem'' (be it a single query or a unified body of general knowledge, as we propose) to a bounded subset of the data in a stream (a window).
  This semantics is henceforth referred to as \emph{global windowing}.
  The iterative execution of individually windowed queries, however, both introduces and \emph{fails to properly consider} the semantics of Complex Event Processing, leading to inconsistencies between the results of identical rule sets expressed in the cited systems \citep{LARS}.

  Truth-maintenance algorithms for continuously changing knowledge bases and logic programs, on the other hand, do support coherent semantics for continuous reasoning.
  However, these semantics cast the problem as that of maintaining a knowledge base and does not support the expression of entailments in streams.
  Without research to extend these semantics to entail concise streams that maintain in their own right the coherence of the existing semantics, these algorithms preclude the further processing of entailments in a stream-processing manner.

  As such, until recently, stream reasoning either lacked a coherent semantics for continuous truth and entailment, or was necessarily a terminus rather than intermediate process in any streamed data pipeline \citep{valle09streamReasoningSummary}.
  However, the work by \citet{LARS} has produced the LARS framework, by which the reasoning semantics of stream processing systems may be expressed and formally analysed.
  LARS extends the semantics of Answer Set Programming (itself an extension of Datalog semantics with features such as Negation as Failure and aggregation) with the notion of streamed data with a point-in-time semantics.
  Paired with this is the notion of a logical windowing operator that supports the expression of windows as part of the general knowledge of a reasoning problem, as is necessary to express reasoning as performed by the iterative execution of continuous queries in the manner already discussed.
  Though this form of windowing also allows the expression of temporally-aware rules in the manner of Complex Event Processing, it does \emph{not} align with or support the expression of \emph{global windowing} derived from stream \emph{querying}.
  Furthermore, the expression of granular windowing such as that expressible in LARS has not been defined in existing Semantic Web reasoning languages such as RIF \citep{w3crif} and OWL \citep{w3cowl2}.
  Finally, the complexity of entailment over programs as defined in LARS makes it unsuitable as a semantics on which to \emph{base} the expression of reasoning in real-time:
  the complexity of entailment in LARS is PSPACE, or co-NP given restrictions on window-nesting, which is generally considered too high a complexity for near-real-time entailment over high-throughput streams \citep{mileo15webSR}.

  % Rete as non-streaming implementationexample (application to dataflow networks)
  Progressing from reasoning semantics to data processing, Rete networks \citep{forgy79} are an example of production systems (a form of rule-based reasoning system) that are particularly suitable for extension with streaming semantics, being modelled as dataflow networks composed of incremental operators.
  These systems send data from operator to operator, with `memories' compiled for two-pass operators in which to retain the specific data received, in much the same way as sliding windows are used in stream processing.
  This makes them functionally similar to many stream-to-stream processing systems, and they have even been a foundation for some such as Sparkwave \citep{sparkwave}.
  However, \citet{forgy79} did not incorporate the semantics of stream processing into the model underlying the Rete algorithm.
  In addition, more recent data-flow-based models either don't consider ``reasoning'' more expressive than RDFS \citep{sparkwave}, or are based on rule/query-level windowing \citep{C-SPARQL,walavalkar08streamingkb}, with problems as already discussed.
  Despite this, data-flow-based models and the vast body of associated literature present a firm foundation for the processing of reasoning tasks in a flexible and scalable manner.

  % overview
  In this paper we present, first, a simple semantics extended from those of positive Datalog that provide a coherent expression of entailments as streams, which we dub Windowed Continuous Datalog.
  The scope of this semantics is to be an expression of inference that extends the static form currently described by Semantic Web reasoning languages with global reasoning semantics, thereby incorporating potentially high-throughput streams of semantic data but resulting in a low enough computational time-complexity that entailments be feasibly computed continuously in real-time.
  As entailment in positive Datalog has polynomial-time complexity and is the equivalent of entailment in RIF-Core \citep{w3crifcore}, which in turn may be used to express the entailment rules of the rule language profile of OWL 2 \citep{w3cowl2profiles}, the continuous semantics of Windowed Continuous Datalog may be used to apply OWL 2 ontologies to streams of RDF data.
  Next, we present the abstract processing model CoDeR that provides for the expression of data-flow processing plans that satisfy the semantics of Windowed Continuous Datalog.
  We also present the system R4 that utilises the Rete pattern-matching algorithm to generate processing plans expressed in the CoDeR model to achieve semantically coherent stream reasoning over sets of rules expressed in RIF-Core.
  Due to the lack of benchmarks for stream \emph{reasoning} use cases, we use single-rule rule-sets based on the individual queries of two stream-\emph{querying} benchmarks to contrast the performance of R4 against that of contemporary systems Etalis and Sparkwave.
  Finally, we discuss differences between the three systems, as well as improvements that could be made to the efficiency of R4 without breaking from the processing model of CoDeR and, therefore, the semantics of Windowed Continuous Datalog.
\end{nestedsection}

>>>>

  \begin{abstract}
    In this paper, we introduce Windowed Continuous Datalog, an extension of
    conventional Datalog that adds the concept of streams of
    temporally-annotated facts to conventional Datalog and defines the
    notion of entailment over a windowed stream. We go on to define CoDeR,
    an abstract model for continuous deductive reasoning that provides
    both a data stream model based on Continuous Datalog and a minimal
    algebra for expressing queries and programs.  Finally, we describe the
    R4 stream reasoner, which implements the CoDeR model as a Rete-based
    dataflow network and provide a comparative evaluation of R4
    against ETALIS and Sparkwave.

@@

\begin{nestedsection}{Introduction}{intro}
  % general framing of the problem (why reason over streams)
  As the Semantic Web becomes more extensive, with more people providing and describing their data in the semantic formats recommended by the W3C, the variety in the nature of the data expressed continues to approach that of previous, more established data formats.
  This includes the growing number of use cases for the continuous processing of streams of semantic data, in which large volumes of time-sensitive data must be processed in near-real time, with the aim of minimising the volume of data transported, duplicated and stored across the Web whilst maximising the variety of problems that can be solved.
  With the focus on reasoning in the Semantic Web stack, varying in expressivity from the low complexity RDF entailment rules through to the high complexity OWL Full, it becomes clear that the greatest challenge in semantic stream processing is that of stream reasoning.

  In the most recent literature, stream reasoning has been characterised as a layer of higher complexity processing on top of Complex Event Processing (CEP) \citep{mileo15webSR}, deeming it suitable only for application to low-throughput macro-event streams.
  However, this expression of reasoning on top of CEP introduces semantics of temporal reasoning \citep{OrgunWadge92,Tuzhilin93}, whereas in many cases \emph{temporally-agnostic} ontologies, which do \emph{not} make use of temporal reasoning, are applicable to \emph{temporally-sensitive} streamed data, which is relevant for an inherently limited amount of time.
  For example, the descriptions of sensors in a sensor network and the various types of reading they produce are often temporally agnostic, while the readings themselves are accurate (and so relevant) for only a short time after they are produced.
  Furthermore, the inferences gleaned from applying the former to the latter are equally temporally sensitive, and so may most accurately be represented as a stream themselves.
  A prime expression of low-complexity reasoning that is capable of such tasks in a static context is the rule language profile of OWL 2 (OWL 2 RL) \citep{w3cowl2profiles}.
  The continuous application of reasoning of this nature to streamed RDF data may obviously be characterised as `stream reasoning', but occupies the same general complexity bracket as stream querying in the hierarchy presented by \citep{mileo15webSR}, making it applicable to high-throughput streams.
  It is this form of stream-based low-complexity reasoning that is the focus of this paper.

  Most stream processing makes use of the concept of \emph{sliding windows} over streams.
  In stream querying, these windows are \emph{global}, in so much as streamed data is stored in windows, then that windowed data \emph{as a whole} is queried \citep{C-SPARQL,EP-SPARQL,CQELS}.
  This may be seen as each datum in a sliding windows being considered to be \emph{true} by a temporally-agnostic continuous query, for the interval from its arrival for the duration of the window.
  It is this notion of windowing that best expresses the form of stream processing we have stated as our focus.

  In contrast, in the recent temporal stream reasoning/CEP literature \citep{LARS}, windows are \emph{localised within} the ``general knowledge'' (e.g. rules in a logic program, the TBox in an ontology), meaning that individual rules/TBox axioms are applied to sliding windows over both streamed data \emph{and} temporally sensitive inferences resulting from other rules/TBox axioms;
  the data is considered to \emph{have been true} at the moment it appeared on a stream/was inferred, and sliding windows retain prior (and future) data according to temporal constraints in the reasoning.
  Localised sliding windows may be used to express a globally windowed \emph{query} (as a single non-recursive rule), and a set of such queries might therefore be used to (partially) express CEP \citep{C-SPARQL,EP-SPARQL,walavalkar08streamingkb,LARS}.

  Localised sliding windows are inherently incapable of expressing temporally-agnostic reasoning over globally windowed data, due to the maintenance of prior inferences within localised sliding windows.
  Inferences from globally windowed streamed data will become untrue as soon as they are no longer implied by the contents of the global window, whereas those inferences would be retained in a localised window for the same duration as any streamed data also retained.
  They may, therefore, be retained \emph{after} the global window would have ceased to imply them, and contribute to further inferences that may, in turn, ``outlive'' them.
  As such, stream reasoning semantics that rely on localised sliding windows (such as that which underpins \citep{LARS}) are unable to express the stream-based reasoning that is the focus of this paper.

  On the other hand, much research has been conducted into the notion of \emph{continuous reasoning}, which has varied from truth-maintenance in ontologies \citep{stream-truth-maintenance}, to the efficient update/deletion of facts \citep{dred,inc-materialisation,ahmeti14rdfsUpdate}, to the addition of time-decaying semantics to Answer Set Programming (ASP) \citep{gebser12streamASP}.
  While these investigate the challenge of changing data in temporally-agnostic (as well as temporal) knowledge-bases, none explicitly provide a semantics in which the entailment of the knowledge base is itself a stream.
  While it is possible to produce streams of data from the changes in a knowledge base, the semantics of such streams have not been explicitly defined or, therefore, thoroughly investigated.

  % motivating scenario
  % Take, for example, the case of extensive sensor networks with many and varied sensors, such as those served by the SemsorGrid4Env project\footnote{\url{http://www.semsorgrid4env.eu/}}.
  % Any stand-alone semantic representation of this data may contain classification information for each sensor, according to what they measure, as well as how frequently and how accurately, as well as other factors.
  % It may also contain explicit data concerning events that are reported by no one sensor, and are instead extrapolated from the base sensor data.
  % In both of these cases, the classification criteria and extrapolation process for macro-events may be expressed as general knowledge that is applicable to all sensor readings.
  % As such, a consumer of the streams produced using SemsorGrid4Env could download the appropriate ontologies once, then use them to reason over a reduced set of streamed data containing only the base readings from the sensor network, thus obtaining the classification and macro-event information desired without having to stream it explicitly.
  % Furthermore, sensor readings are often superseded by later readings of the same type from the same sensor, and so older readings may be discarded as time progresses.

  % % *brief* summary of state of the art (to be expanded in related work)
  % Recent research has approached the task of stream reasoning from two different directions, though each is flawed with respect to .
  % %   RDF stream processing and reasoning
  % The RDF stream processing community has proposed extensions to their continuous semantic query languages to support the \emph{iterative/recursive execution} of queries over the results of other queries \citep{C-SPARQL,EP-SPARQL,walavalkar08streamingkb}, in which each query expresses a rule in a logic program.
  % This reflects precisely the approach to reasoning using Datalog in deductive databases, in which queries may be executed on intensional relations whose definitions must be used in turn to query the database, recursing in this manner until base only extensional relations are queried.
  % Meanwhile, the reasoning community has explored the concept of continuously reasoning over streamed data in a variety of forms, primarily through a range of truth maintenance algorithms \citep{inc-materialisation,stream-truth-maintenance}.
  % %%%FIXME
  % NEED SOMETHING ON TIME-DECAY MODEL

  % %   issues with existing art: no explicit semantics for reasoning
  % While the iterative execution of queries as rules is sufficient for performing reasoning to the expressivity of Datalog when applying one-shot queries to static data, as practised in deductive databases, the query-by-query windowing semantics of most stream-querying systems do not translate to a semantics for multi-rule stream reasoning of the kind investigated in this paper.
  % This is because the windowing semantics proposed for stream querying involve applying the entirety of a ``problem'' (be it a single query or a unified body of general knowledge, as we propose) to a bounded subset of the data in a stream (a window).
  % This semantics is henceforth referred to as \emph{global windowing}.
  % The iterative execution of individually windowed queries, however, both introduces and \emph{fails to properly consider} the semantics of Complex Event Processing, leading to inconsistencies between the results of identical rule sets expressed in the cited systems \citep{LARS}.

  % Truth-maintenance algorithms for continuously changing knowledge bases and logic programs, on the other hand, do support coherent semantics for continuous reasoning.
  % However, these semantics cast the problem as that of maintaining a knowledge base and does not support the expression of entailments in streams.
  % Without research to extend these semantics to entail concise streams that maintain in their own right the coherence of the existing semantics, these algorithms preclude the further processing of entailments in a stream-processing manner.

  % As such, until recently, stream reasoning either lacked a coherent semantics for continuous truth and entailment, or was necessarily a terminus rather than intermediate process in any streamed data pipeline \citep{valle09streamReasoningSummary}.
  % However, the work by \citet{LARS} has produced the LARS framework, by which the reasoning semantics of existing stream processing systems may be expressed and formally contrasted.
  % LARS extends the semantics of Answer Set Programming (itself an extension of Datalog semantics with features such as Negation as Failure and aggregation) with the notion of streamed data with a point-in-time semantics.
  % Paired with this is the notion of a logical windowing operator that supports the expression of windows as part of the general knowledge of a reasoning problem, as is necessary to express reasoning as performed by the iterative execution of continuous queries in the manner already discussed.
  % Though this form of windowing also allows the expression of temporally-aware rules in the manner of Complex Event Processing, it does \emph{not} align with or support the expression of \emph{global windowing} derived from stream \emph{querying}.
  % Furthermore, a standard for the expression of \emph{local} windowing such as that expressible in LARS has not been defined in existing Semantic Web reasoning languages such as RIF \citep{w3crif} and OWL \citep{w3cowl2}.
  % Finally, the complexity of entailment over programs as defined in LARS makes it unsuitable as a semantics on which to \emph{base} the expression of reasoning in real-time:
  % the complexity of entailment in LARS is PSPACE, or co-NP given restrictions on window-nesting, which is generally considered too high a complexity for near-real-time entailment over high-throughput streams \citep{mileo15webSR}.

  % Rete as non-streaming implementationexample (application to dataflow networks)
  Progressing from stream-based reasoning semantics to data processing, Rete networks \citep{forgy79} are an example of production systems (a form of rule-based reasoning system) that are particularly suitable for extension with streaming semantics, being modelled as dataflow networks composed of incremental operators.
  These systems send data from operator to operator, with `memories' compiled for two-pass operators in which to retain the specific data received, in much the same way as sliding windows are used in stream processing.
  This makes them functionally similar to many stream-to-stream processing systems, and they have even been a foundation for some such as Sparkwave \citep{sparkwave}.
  However, \citet{forgy79} did not incorporate the semantics of stream processing into the model underlying the Rete algorithm.
  In addition, more recent data-flow-based models either don't consider ``reasoning'' more expressive than RDFS \citep{sparkwave}, or are based on rule/query-level windowing \citep{C-SPARQL,walavalkar08streamingkb}, which has already been shown to be incompatible with globally-windowed stream-based reasoning.
  Despite this, data-flow-based models and the vast body of associated literature present a firm foundation for the processing of stream-based reasoning tasks in a flexible and scalable manner.

  % overview
  In this paper we present, first, a simple semantics that extends those of positive Datalog to both incorporate streamed data and provide an expression of entailments over time as streams.
  The scope of this semantics, which we dub Windowed Continuous Datalog, is to be an expression of inference that extends the static form currently described by Semantic Web reasoning languages with global windowing semantics, thereby incorporating potentially high-throughput streams of semantic data while retaining a low enough computational time-complexity that streams of entailments may feasibly be computed continuously in real-time.
  As entailment in positive Datalog has polynomial-time complexity and is the equivalent of entailment in RIF-Core \citep{w3crifcore}, which in turn may be used to express the entailment rules of the rule language profile of OWL 2 \citep{w3cowl2profiles}, the continuous semantics of Windowed Continuous Datalog may be used to apply OWL 2 ontologies to streams of RDF data.

  Next, we present the abstract processing model CoDeR that provides for the expression of data-flow processing plans that satisfy the semantics of Windowed Continuous RIF-Core.
  We also present the system R4 that utilises the Rete pattern-matching algorithm to generate processing plans expressed in the CoDeR model to achieve semantically sound stream-based reasoning over sets of rules expressed in RIF-Core.
  Due to the lack of benchmarks for expressive stream-based \emph{reasoning} use cases, we use the modified big-data/RDFS evaluation described by \citet{sparkwave} to contrast the performance of R4 against that of contemporary systems Etalis \citep{EP-SPARQL} and Sparkwave \citep{sparkwave}.
  From this, we assert that R4 is empirically faster than these contemporaries, despite having a more expressive underlying semantics.
  Finally, we briefly discuss possible extensions to R4 to further improve its quality of service characteristics.
\end{nestedsection}
